Simulating interventions to prevent gastro-intestinal transmission of _Clostridioides difficile_ and antimicrobial-resistant bacteria in healthcare settings: a model calibration study

### 2025-12-30 paper/paper.qmd
round of edits

---
title: Gastro-intestinal transmission of _Clostridioides difficile_ and antimicrobial-resistant bacteria in healthcare settings -- calibration of a simulation model
authors:
  - name: Alessandro Gerada
    affiliation: University of Liverpool
    roles: writing
    corresponding: true
bibliography: references.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/refs/heads/master/elsevier-vancouver.csl
filters:
  - acronyms
acronyms:
  keys:
  - shortname: CDI
    longname: _Clostridium difficile_ infection
    markdown: longname
  - shortname: VRE
    longname: vancomycin-resistant _Enterococcus_
    markdown: longname
  - shortname: ESBL
    longname: extended-spectrum beta-lactamase
  - shortname: CPE
    longname: carbapenemase-producing Enterobacterales
  - shortname: WHO
    longname: World Health Organization
  - shortname: IPC
    longname: infection prevention and control
  - shortname: ABM
    longname: agent-based model
  - shortname: MRSA
    longname: methicillin-resistant _Staphylococcus aureus_
    markdown: longname
  - shortname: SARS-CoV-2
    longname: severe acute respiratory syndrome coronavirus 2
  - shortname: ABC
    longname: approximate Bayesian computation
  - shortname: IQR
    longname: interquartile range
  - shortname: FIFO
    longname: first-in, first-out
  - shortname: SD
    longname: system dynamics
  insert_loa: false
  insert_links: false
---

@@

\acr[case=sentence, first_use]{IPC} measures are designed to prevent the spread of these pathogens within healthcare settings. Commonly used interventions include hand hygiene, environmental cleaning, and the isolation of infected patients [@worldhealthorganizationGlobalReportInfection2022].
A key role of \acr{IPC} teams is to prepare for, declare, investigate, and control outbreaks of healthcare-associated infections.
In practice, an outbreak is declared after rates of infection have been compared to historical rates, and this is followed by a series of interventions to control the outbreak.
These interventions may include increased frequency and thoroughness of environmental cleaning, cohorting of infected patients (particularly into single-occupancy _en suite_ rooms), and increased hand hygiene compliance.
The impact of these interventions is often difficult to quantify.
Although evidence suggests that the interventions can help control outbreaks, studies almost always report their findings after a suite of interventions have been implemented -- therefore, the relative importance of each intervention is not usually quantifiable.
Furthermore, studies report findings from a population that may be very different from the outbreak population being managed -- differences can range from patient factors such as co-morbidities and rates of antimicrobial resistance, to institution practices such as rates of antimicrobial usage, and to structural differences such as hospital layout.

@@

Since the simulation's target diseases were associated with long term colonisation and prolonged risk of transmission to other patients, it was assumed that infection was not reversible during the simulation, i.e., that infected patients remained infected until discharge. At each time increment, the probability of a non-infected patient $i$ acquiring infection (i.e., transitioning from a non-infected to infected state)---$P_i(t)$---depended on: 1) a baseline risk of *de novo* infection; and 2) the usage of contaminated toilets. The probability of infection through these two routes---$P_{\text{de novo}, i}(t)$ and $P_{\text{toilet}, i}(t)$, respectively---were independent; therefore, the probability of patient $i$ acquiring infection at time $t$ was given by:

@@

After calibrating the \acr{ABM} using \acr{ABC}, three groups of parameters were produced: 1) baseline parameters, which aimed to generate similar rates of infection to the basal rate (i.e., pre- and post-outbreak); 2) outbreak parameters, which led to the increased transmission observed during the outbreak period; and 3) control parameters, which reflect enhanced infection control practice that led to outbreak control. A summary of the posterior estimates of the parameters (reported as the mean and standard deviation of the parameters from the rejection \acr{ABC}) is shown in @tbl-posteriors. The densities of the posterior parameter distributions are shown in @fig-posterior-densities. The posterior distributions of the parameters were consistent with the expected changes in the parameters during the outbreak and control periods (e.g., the toilet cleaning rate changed from `r b_toilet_clean_rate` to `r o_toilet_clean_rate`, and to `r c_toilet_clean_rate` during the baseline, outbreak, and control periods respectively).

```{r tbl-posteriors, include=TRUE}
#| tbl-cap: "Summary of parameter posteriors after calibration using approximate Bayesian computation."
#| echo: false
params_table <- abc_params_outbreak_control$unadj.values %>% 
  as_tibble() %>% 
  mutate("community-colonisation-rate" = `community-colonisation-rate` * 100,
         "o-community-colonisation-rate" = `o-community-colonisation-rate` * 100,
         "o-antibiotic-prescription-rate" = `o-antibiotic-prescription-rate` * 100,
         "c-antibiotic-prescription-rate" = `c-antibiotic-prescription-rate` * 100) %>% 
  dplyr::select(!any_of(c("outbreak-start", "outbreak-end", "control-end"))) %>%
  dplyr::rename_with(.fn = \(.x) paste0("b-", .x), .cols = !starts_with("o-") & !starts_with("c-")) %>% 
  summarise(across(everything(), ~ paste0(
    round(mean(.x), 2),
    " (",
    round(sd(.x), 2),
    ")"
    )))

@@

param_names <- c("toilet-frequenting-rate" = "Toilet frequenting rate",
                 "toilet-cleaning-rate" = "Toilet cleaning rate",
                 "toilet-contamination-effect" = "Toilet contamination effect",
                 "toilet-cleaning-effect" = "Toilet cleaning effect",
                 "community-colonisation-rate" = "Community infection rate",
                 "antibiotic-prescription-rate" = "Antibiotic prescription rate",
                 "antibiotic-effect" = "Antibiotic effect",
                 "random-colonisation" = "Random infection",
                 "side-room-delay-mean" = "Side room delay mean")

params_table %>% 
  pivot_longer(cols = everything(), names_to = c("period", "param"), values_to = "value",
               names_pattern = "(b|o|c)-(.*)") %>% 
  pivot_wider(names_from = period, values_from = value) %>% 
  # reorder cols
  dplyr::select(param, b, o, c) %>% 
  dplyr::rename(baseline = b, outbreak = o, control = c) %>% 
  mutate(units = param) %>% 
  relocate(units, .after = param) %>%
  flextable() %>% 
  labelizor(j = "units", part = "body",
            labels = param_units) %>% 
  labelizor(j = "param", part = "body",
            labels = param_names) %>% 
  width(width = 1) %>% 
  font(fontname="Times New Roman",
       part = "all") %>% 
  fontsize(size = 10, part = "all")
```

@@

Although \acrs{ABM} have been shown to be useful for simulating and modelling infectious disease transmission, they have limitations. The model described here is a simplification of what is likely to be a highly intricate and complicated system. For example, transmission was assumed to be 'random' or acquired through the use of a contaminated toilet. In reality, other transmission pathways likely exist, such as through contact with staff or contaminated bedding. Antimicrobial exposure was considered to be a binary risk factor for an agent becoming infected. Antimicrobial class and duration of exposure are factors that are not accounted for in this model. An \acr{ABM} that captures all the details that influence the spread of infection in a hospital setting is unlikely to be feasible, due to the computational challenge and the large number of parameters that would need to be estimated. A reasonable approach is for the model to be as simple as possible, while still capturing the features required to answer the research question [@sunSimpleComplicatedAgentbased2016]. Another limitation of \acrs{ABM} is the requirement for detailed data to calibrate and validate the model. Here, the data used was primarily from a retrospective descriptive dataset. As such, information on certain parameters was not available (e.g., the proportion of patients that were administered an antimicrobial during their admission) -- these parameters were estimated from the literature. An important future direction would be to prospectively inform and validate the model with data collected specifically for this purpose.

>>>>

---
title: "Simulating interventions to prevent gastro-intestinal transmission of _Clostridioides difficile_ and antimicrobial-resistant bacteria in healthcare settings: a model calibration study"
bibliography: references.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/refs/heads/master/elsevier-vancouver.csl
filters:
  - acronyms
  - authors-block
acronyms:
  keys:
  - shortname: CDI
    longname: _Clostridium difficile_ infection
    markdown: longname
  - shortname: VRE
    longname: vancomycin-resistant _Enterococcus_
    markdown: longname
  - shortname: ESBL
    longname: extended-spectrum beta-lactamase
  - shortname: CPE
    longname: carbapenemase-producing Enterobacterales
  - shortname: WHO
    longname: World Health Organization
  - shortname: IPC
    longname: infection prevention and control
  - shortname: ABM
    longname: agent-based model
  - shortname: MRSA
    longname: methicillin-resistant _Staphylococcus aureus_
    markdown: longname
  - shortname: SARS-CoV-2
    longname: severe acute respiratory syndrome coronavirus 2
  - shortname: ABC
    longname: approximate Bayesian computation
  - shortname: IQR
    longname: interquartile range
  - shortname: FIFO
    longname: first-in, first-out
  - shortname: SD
    longname: system dynamics
  insert_loa: false
  insert_links: false
authors:
  - name: Alessandro Gerada
    affiliations:
      - ref: APT
      - ref: LCL
    orcid: 0000-0002-6743-4271
    email: alessandro.gerada@liverpool.ac.uk
    corresponding: true
  - name: Charlotte Brookefield
    affiliations:
      - ref: APT
      - ref: LCL
  - name: Alex Howard
    affiliations:
      - ref: APT
      - ref: LCL
  - name: Nada Reza
    affiliations:
      - ref: APT
  - name: Conor Rosato
    affiliations:
      - ref: APT
  - name: Peter L. Green
    affiliations:
      - ref: APT
      - ref: SOE
  - name: William Hope
    affiliations:
      - ref: APT
      - ref: LCL
affiliations:
  - id: APT
    name: Antimicrobial Pharmacodynamics and Therapeutics, Department of Clinical Pharmacology and Therapeutics, Institute of Systems, Molecular & Integrative Biology, University of Liverpool
  - id: LCL
    name: Department of Infection and Immunity, Liverpool Clinical Laboratories, Clinical Support Services Building (CSSB), Liverpool University Hospitals NHS Foundation Trust
  - id: SOE
    name: School of Engineering, Foundation Building, Brownlow Hill, University of Liverpool, Liverpool, L69 7ZX
---

@@

\acr[case=sentence, first_use]{IPC} measures are designed to prevent the spread of these pathogens within healthcare settings. Commonly used interventions include hand hygiene, environmental cleaning, and the isolation of infected patients [@worldhealthorganizationGlobalReportInfection2022].
A key role of \acr{IPC} teams is to prepare for, declare, investigate, and control outbreaks of healthcare-associated infections.
In practice, an outbreak is declared after rates of infection have been compared to historical rates, and this is followed by a series of interventions to control the outbreak.
These interventions may include increased frequency and thoroughness of environmental cleaning, cohorting of infected patients (particularly into single-occupancy _en suite_ rooms), and increased hand hygiene compliance.
The impact of these interventions is often difficult to quantify.
Although evidence suggests that the interventions can help control outbreaks, studies almost always report their findings after a suite of interventions have been implemented -- therefore, the relative importance of each intervention is not usually quantifiable [@baierMultimodalInfectionControl2018; @davisImpactPreventionBundle2016].
Furthermore, studies report findings from a population that may be very different from the outbreak population being managed -- differences can range from patient factors such as co-morbidities and rates of antimicrobial resistance, to institution practices such as rates of antimicrobial usage, and to structural differences such as hospital layout.

@@

Since the simulation's target diseases were associated with long term colonisation and prolonged risk of transmission to other patients, it was assumed that infection was not reversible during the simulation, i.e., that infected patients remained infected until discharge. At each time increment, the probability of a non-infected patient $i$ acquiring infection (i.e., transitioning from a non-infected to infected state)---$P_i(t)$---depended on: 1) a baseline risk of *de novo* infection; and 2) the usage of contaminated toilets. In this context, *de novo* infection was assumed to capture routes of transmission outwith toilet use, and infection which was not directly related to a traceable transmission event. The probability of infection through these two routes---$P_{\text{de novo}, i}(t)$ and $P_{\text{toilet}, i}(t)$, respectively---were independent; therefore, the probability of patient $i$ acquiring infection at time $t$ was given by:

@@

After calibrating the \acr{ABM} using \acr{ABC}, three groups of parameters were produced: 1) baseline parameters, which aimed to generate similar rates of infection to the basal rate (i.e., pre- and post-outbreak); 2) outbreak parameters, which led to the increased transmission observed during the outbreak period; and 3) control parameters, which reflect enhanced infection control practice that led to outbreak control. A summary of the posterior estimates of the parameters (reported as the median and interquartile range of the parameters from the rejection \acr{ABC}) is shown in @tbl-posteriors. The densities of the posterior parameter distributions are shown in @fig-posterior-densities. The posterior distributions of the parameters were consistent with the expected changes in the parameters during the outbreak and control periods (e.g., the toilet cleaning rate changed from `r b_toilet_clean_rate` to `r o_toilet_clean_rate`, and to `r c_toilet_clean_rate` during the baseline, outbreak, and control periods respectively).

```{r tbl-posteriors, include=TRUE}
#| tbl-cap: "Summary of parameter posteriors after calibration using approximate Bayesian computation."
#| echo: false
params_table <- abc_params_outbreak_control$unadj.values %>% 
  as_tibble() %>% 
  mutate("community-colonisation-rate" = `community-colonisation-rate` * 100,
         "o-community-colonisation-rate" = `o-community-colonisation-rate` * 100,
         "o-antibiotic-prescription-rate" = `o-antibiotic-prescription-rate` * 100,
         "c-antibiotic-prescription-rate" = `c-antibiotic-prescription-rate` * 100) %>% 
  dplyr::select(!any_of(c("outbreak-start", "outbreak-end", "control-end"))) %>%
  dplyr::rename_with(.fn = \(.x) paste0("b-", .x), .cols = !starts_with("o-") & !starts_with("c-")) %>% 
  summarise(across(everything(), ~ {
    med <- median(.x)
    q25 <- quantile(.x, 0.25)
    q75 <- quantile(.x, 0.75)
    paste0(round(med, 2), " (", round(q25, 2), "–", round(q75, 2), ")")
  }))

@@

param_names <- c("toilet-frequenting-rate" = "Patient toilet frequenting rate",
                 "toilet-cleaning-rate" = "Toilet cleaning rate",
                 "toilet-contamination-effect" = "Toilet contamination effect",
                 "toilet-cleaning-effect" = "Toilet cleaning effect",
                 "community-colonisation-rate" = "Community infection rate",
                 "antibiotic-prescription-rate" = "Antibiotic prescription rate",
                 "antibiotic-effect" = "Antibiotic effect",
                 "random-colonisation" = "Random infection",
                 "side-room-delay-mean" = "Side room delay mean")

params_table %>% 
  pivot_longer(cols = everything(), names_to = c("period", "param"), values_to = "value",
               names_pattern = "(b|o|c)-(.*)") %>% 
  pivot_wider(names_from = period, values_from = value) %>% 
  # reorder cols
  dplyr::select(param, b, o, c) %>% 
  dplyr::rename(baseline = b, outbreak = o, control = c) %>% 
  mutate(units = param) %>% 
  relocate(units, .after = param) %>%
  flextable() %>% 
  add_header_row(values = c("Parameter name", "Units", "Value"), colwidths = c(1, 1, 3)) %>% 
  set_header_labels(param = "", units = "", baseline = "Baseline", outbreak = "Outbreak", control = "Control") %>% 
  align(i = 1, j = 3, align = "center", part = "header") %>% 
  labelizor(j = "units", part = "body",
            labels = param_units) %>% 
  labelizor(j = "param", part = "body",
            labels = param_names) %>% 
  width(width = 1) %>% 
  font(fontname="Times New Roman",
       part = "all") %>% 
  fontsize(size = 10, part = "all")
```

@@

Although \acrs{ABM} have been shown to be useful for simulating and modelling infectious disease transmission, they have limitations. The model described here is a simplification of what is likely to be a highly intricate and complicated system. For example, transmission was assumed to be 'random' or acquired through the use of a contaminated toilet. In reality, other transmission pathways likely exist, such as through contact with staff or contaminated bedding. Antimicrobial exposure was considered to be a binary risk factor for an agent becoming infected. Antimicrobial class and duration of exposure are factors that are not accounted for in this model. An \acr{ABM} that captures all the details that influence the spread of infection in a hospital setting is unlikely to be feasible, due to the computational challenge and the large number of parameters that would need to be estimated. A reasonable approach is for the model to be as simple as possible, while still capturing the features required to answer the research question [@sunSimpleComplicatedAgentbased2016]. Another limitation of \acrs{ABM} is the requirement for detailed data to calibrate and validate the model. Here, the data used were primarily from a retrospective descriptive dataset. As such, information on certain parameters was not available (e.g., the proportion of patients that were administered an antimicrobial during their admission) -- these parameters were estimated from the literature. An important future direction would be to prospectively inform and validate the model with data collected specifically for this purpose.
